Vagueness, Noise, and Communication

### 2014-09-28 paper/03_simulations.tex
results report

Now that we understand better what diffusion does, let us look at how
diffusion interacts with optimization of behavior, as described by the
replicator dynamics.

@@

The main question we should address is whether diffusion and
fitness-based replication reasonably interact, and if so, whether the
inclusion of diffusion has any noteworthy effects on the evolving
meaning of signals. Clearly, if $\impairment = 0$, the \rdd reduces to
the \rd. If $\impairment \rightarrow \infty$, the diffusion component
takes over and the \rdd reduces to the trivial iterated diffusion that
we looked at in the previous section. For reasonable, in-between
levels of imprecision $\impairment$, the \rdd leads to communicatively
successful, yet vague signal meaning. Moreover, suitable levels of
imprecision can have further accelerating and, therefore, unifying
effects on meaning evolution.

To explore the \rdd, we looked at numerical simulations in sim-max
games with two signals whose states were evenly spaced elements of the
unit interval. The state space always included 0 and 1. We assumed
that priors were flat: $\Pr(\state) = \Pr(\state')$ for all $\state,
\state'$. As for utility functions, we implemented another
Nosofsky-style function:
\begin{align}
  \label{eq:NosofskyUtil}
  \util(\mystate{i}, \mystate{j} \myts ; \myts \toler) =
      \begin{cases}
    1 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} = \mystate{j} \\
    0 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} \neq \mystate{j} \\
 \expo \left ( -  \frac{\card{\mystate{i} - \mystate{j}}^s}{ \toler^2} \right ) \,,& \textrm{otherwise} \\
    \end{cases}
\end{align}
The tolerance parameter $\toler$ models the amount of tolerable
pragmatic slack or communicative imprecision. For This choice of
utility function is governed partly by convenience in parallel to the
similarity function, but also because it has, as we believe, the right
general properties for a communicative payoff function. Unlike
utilities that are, say, linearly or quadratically decreasing
\citep[as used
by][]{JagerMetzger2011:Voronoi-Languag,FrankeJager2010:Vagueness-Signa},
utilities that are exponentially decreasing in negative quadratic
distance can model situations where a small amount of imprecision in
communication is tolerable, whereas similarly small differences in
intolerably off interpretations matter very little, with a smooth
transition between these regimes.

\begin{figure}
  \centering
  \subfloat[][$\ns = 90$, $\toler = 0.2$, $\impairment = 0.1$]{
    \includegraphics[width = 0.5\textwidth]{plots/strat_example_NS-90_tol-2_imp1_ind3501.pdf}
  }
  \subfloat[][$\ns = 90$, $\toler = 0.05$, $\impairment = 0$]{
    \includegraphics[width = 0.5\textwidth]{plots/strat_example_NS-90_tol-005_imp0_ind3002.pdf}
  }

  \subfloat[][$\ns = 10$, $\toler = 0.05$, $\impairment = 0$]{
    \includegraphics[width = 0.5\textwidth]{plots/strat_example_NS-10_tol-005_imp0_ind1001.pdf}
  }
  \subfloat[][$\ns = 6$, $\toler = 0.05$, $\impairment = 0.05$]{
    \includegraphics[width = 0.5\textwidth]{plots/strat_example_NS-10_tol-005_imp005_ind1201.pdf}
  }
  \caption{Example strategies under \rdd at stopping time of our simulations.}
  \label{fig:example_strats}
\end{figure}

We ran 50 trials of the \rdd, starting with randomly samples sender
and receiver strategies, for each triplet of independent parameter
values: $\ns = \card{T} \in \set{6, 10, 50, 90}$, $\impairment \in
\set{0, 0.05, 0.1, 0.2, 0.3}$, $\toler \in \set{0.05, 0.1, 0.2,
  0.3}$. Each trial ran for a maximum of 200 update steps of the
\rdd. This may not be enough to guarantee convergence to the eventual
attracting state, but we were interested especially in whether the
dynamics could lead to reasonable outcomes in reasonable time. A trial
was considered converged if the total amount of change between
strategies before and after the current \rdd step was smaller than
0.001. Examples for the resulting strategy pairs at stopping time are
given in Figure~\ref{fig:example_strats}.

\begin{itemize}
\item we see some interesting stuff in these strategies
\item now we look at the whole bunch
\item introduce some measures
\end{itemize}

\newpage

In order to study the long-term behavior of the noise-perturbed sim-max games, we conducted computer simulations for different combinations of initial parameters.
Namely, we were interested in investigating the impact of different values of impairment and its relation with the size of the state space.
\mytodo{JPC}{Justify why size of state space is interesting}

\subsubsection{Metrics}
\label{sec:metrics}
To be able to quantitively analyze the results of our simulations, we define a number of metrics to capture different characteristics of sim-max languages.

\paragraph{Entropy.}
This classic information-theoretic metric captures the amount of uncertainty in a signaling strategy.
The most natural way to define it is in terms of mixed strategies rather than behavioral strategies.
Let us recall that mixed sender (receiver) strategies are functions $\Smixed \in \Delta(\Messgs^\States)$ ($\Rmixed \in (\Acts^\Messgs)$).
We can define the entropy of a sender strategy as
\begin{align*}
  E(\Smixed) = \sum_{\Spure \in \Messgs^\States} \Smixed(\Spure) \cdot \log(\Smixed(\Spure))
\end{align*} 
and the entropy of a receiver strategy as
\begin{align*}
  E(\Rmixed) = \sum_{\Rpure \in \Acts^\Messgs} \Rmixed(\Rpure) \cdot \log(\Rmixed(\Rpure)) \,.
\end{align*} 
These metrics are computationally expensive to calculate, since the size of the domain over which the sum is computed grows exponentially with the number of choice points.
Therefore, we converted\footnote{See Appendix~\ref{sec:proofs}.} the above definitions into equivalent metrics defined in terms of behavioral strategies.
These are as follows:
\begin{align*}
  E(\Sstrat) = -\sum_{\state \in \States} \sum_{\messg \in \Messgs} \Sstrat(\messg \probbar \state) \cdot \log(\Sstrat(\messg \probbar \state))
\end{align*} 
\begin{align*}
  E(\Rstrat) = -\sum_{\messg \in \Messgs} \sum_{\act \in \Acts} \Rstrat(\act \probbar \messg) \cdot \log(\Rstrat(\act \probbar \messg)) \,.
\end{align*}
The metrics are lower bounded by $0$ and upper bounded by, respectively, $\log(|\Messgs^\States|) = |\States| \cdot \log(|\Messgs|)$ and $\log(|\Acts^\Messgs|) = |\Messgs| \cdot \log(|\Acts|)$, thus we can normalize them by dividing by these values.

\paragraph{Informativity.}
\mytodo{JPC}{Informativity is perhaps a bad name for the receiver...}
The quantity of information in a signal is an old notion that also goes back to the start of information theory.
Skyrms~\citet[ch.~3]{Skyrms2010:Signals} discusses its use in the context of signaling games.
The main idea is that we can quantify the amount of information about a state $\state$ in a message $\messg$ via the relation between the probability of the state given the message $\Pr(\state \,|\, \messg)$ and the unconditional probability of the state $\Pr(\state)$.
Following Bayes' theorem, we can express $\Pr(\state \,|\, \messg)$ as $\frac{\Pr(\state) \cdot \Pr(\messg \,|\, \state)}{\Pr(\messg)}$.
We then have $\Pr(\messg \,|\, \state) = \Sstrat(\state,\messg)$ and $\Pr(\messg) = \sum_{\state^\prime \in \States} \Pr(\state^\prime) \cdot \Sstrat(\state^\prime,\messg)$.
Finally, we equate sender infomativity with the average overall information about states in each signal.
Based on the definition by Skyrms~\citet[p.~36]{Skyrms2010:Signals}, and the considerations above, we define sender informativity as follows:
\begin{align*}
  I(\Sstrat) = \frac{1}{|\Messgs|} \sum_{\messg \in \Messgs} \sum_{\state \in \States} \frac{\Pr(\state) \cdot \Sstrat(\state,\messg)}{\sum_{\state^\prime \in \States} \Pr(\state^\prime) \cdot \Sstrat(\state^\prime,\messg)} \cdot \log \left(\frac{\Sstrat(\state,\messg)}{\sum_{\state^\prime \in \States} \Pr(\state^\prime) \cdot \Sstrat(\state^\prime,\messg)} \right) \,.
\end{align*}
Conversely, we can quantify the amount of information about an act in a message.
We equate receiver informativity with the average overall information about acts in each signal.
Based on Skyrms' definition~\citet[p.~39]{Skyrms2010:Signals}, we define receiver informativity as follows:
\begin{align*}
  I(\Sstrat,\Rstrat) = \frac{1}{|\Messgs|} \sum_{\messg \in \Messgs} \sum_{\act \in \Acts} \Rstrat(\messg,\act) \cdot \log \left(\frac{\Rstrat(\messg,\act)}{\sum_{\state \in \States} \Pr(\state) \cdot \sum_{\messg^\prime \in \Messgs} \Sstrat(\state,\messg^\prime) \cdot \Rstrat(\messg^\prime,\act)} \right) \,.
\end{align*}
Both metrics are bounded between $0$ and $1$.

@@

We run the noise-perturbed replicator dynamics with random initial strategies $\Sstrat_0$ and $\Rstrat_0$ until a convergence criterion is met, namely that the total absolute change in both sender and receiver strategies does not amount to more than $0.01$, \emph{i.e.}:
\begin{align*}
  \sum_{\state \in \States} \sum_{\messg \in \Messgs} \left| \Sstrat^\prime(\state,\messg) - \Sstrat(\state,\messg) \right| < 0.01
\end{align*}
and
\begin{align*}
  \sum_{\messg \in \Messgs} \sum_{\state \in \States} \left| \Rstrat^\prime(\messg,\state) - \Rstrat(\messg,\state) \right| < 0.01 \,.
\end{align*}

>>>>

@@

Now that we understand better what diffusion does, let us look at how
diffusion interacts with optimization of behavior, as described by the
replicator dynamics. The main question we should address is whether
diffusion and fitness-based replication reasonably interact, and if
so, whether the inclusion of diffusion has any noteworthy effects on
the evolving meaning of signals. Clearly, if $\impairment = 0$, the
\rdd reduces to the \rd. If $\impairment \rightarrow \infty$, the
diffusion component takes over and the \rdd reduces to the trivial
iterated diffusion that we looked at in the previous section. For
reasonable in-between levels of imprecision $\impairment$, the \rdd
leads to communicatively successful, yet vague signal
meaning. Moreover, suitable levels of imprecision can have further
accelerating and, therefore, unifying effects on meaning evolution.

To explore the \rdd, we looked at numerical simulations in sim-max
games with two signals whose states were evenly spaced elements of the
unit interval that always included 0 and 1. We assumed that priors
were flat: $\Pr(\state) = \Pr(\state')$ for all $\state, \state'$. As
for utility functions, we implemented another Nosofsky-style function:
\begin{align}
  \label{eq:NosofskyUtil}
  \util(\mystate{i}, \mystate{j} \myts ; \myts \toler) =
      \begin{cases}
    1 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} = \mystate{j} \\
    0 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} \neq \mystate{j} \\
 \expo \left ( -  \frac{\card{\mystate{i} - \mystate{j}}^s}{ \toler^2} \right ) \,,& \textrm{otherwise} \\
    \end{cases}
\end{align}
The tolerance parameter $\toler$ models the amount of tolerable
pragmatic slack or communicative imprecision. This choice of utility
function is governed partly by convenience in parallel to the
similarity function, but also because it has, as we believe, the right
general properties for a communicative payoff function. Unlike
utilities that are, say, linearly or quadratically decreasing
\citep[as used
by][]{JagerMetzger2011:Voronoi-Languag,FrankeJager2010:Vagueness-Signa},
utilities that are exponentially decreasing in negative quadratic
distance can model situations where a small amount of imprecision in
communication is tolerable, whereas similarly small differences in
intolerably off interpretations matter very little, with a smooth
transition between these regimes \citep{OConnor2013:The-Evolution-o}.

  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/strat_example_NS-90_tol-01_imp0_ind3098.pdf}
    \caption{$\ns = 90$, $\toler = 0.1$, $\impairment = 0$}
    \label{fig:example_stratsA}
  \end{subfigure}
  \hfill
  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/strat_example_NS-90_tol-01_imp01_ind3452.pdf}
    \caption{$\ns = 90$, $\toler = 0.1$, $\impairment = 0.1$}
    \label{fig:example_stratsB}
  \end{subfigure}

  % \begin{subfigure}[]{0.45\textwidth}
  %   \includegraphics[width=\textwidth]{plots/strat_example_NS-10_tol-005_imp0_ind1001.pdf}
  %   \caption{$\ns = 10$, $\toler = 0.1$, $\impairment = 0$}
  %   \label{fig:example_stratsC}
  % \end{subfigure}
  % \hfill
  % \begin{subfigure}[]{0.45\textwidth}
  %   \includegraphics[width=\textwidth]{plots/strat_example_NS-10_tol-005_imp005_ind1201.pdf}
  %   \caption{$\ns = 10$, $\toler = 0.1$, $\impairment = 0.05$}
  %   \label{fig:example_stratsD}
  % \end{subfigure}

  \caption{Example strategies under \rdd at stopping time of our simulations.}
  \label{fig:example_strats}
\end{figure}

We ran 50 trials of the \rdd, starting with randomly sampled sender
and receiver strategies, for each triplet of independent parameter
values: $\ns = \card{T} \in \set{6, 10, 50, 90}$, $\impairment \in
\set{0, 0.05, 0.1, 0.2, 0.3}$, $\toler \in \set{0.05, 0.1, 0.2,
  0.3}$. Each trial ran for a maximum of 200 update steps of the
\rdd. This may not be enough to guarantee convergence to the eventual
attracting state, but we were interested especially in whether the
dynamics could lead to reasonable outcomes in reasonable time. A trial
was considered converged and stopped before the maximum of 200 rounds
if the total amount of change between strategies before and after the
current \rdd step was smaller than 0.001, i.e., if both:\todo{say what
the prime means}
\begin{align*}
  \sum_{\state \in \States} \sum_{\messg \in \Messgs} \left|
    \Sstrat^\prime(\state,\messg) - \Sstrat(\state,\messg) \right| <
  0.001 & \text{\ \ and} &
  \sum_{\messg \in \Messgs} \sum_{\state \in \States} \left| \Rstrat^\prime(\messg,\state) - \Rstrat(\messg,\state) \right| < 0.001 \,.
\end{align*}
Representative examples for the resulting strategy pairs at stopping
time are given in Figure~\ref{fig:example_strats}.

Figure~\ref{fig:example_stratsA} shows a strategy pair at stopping
time with 90 states, tolerance $\toler = 0.1$ and imprecision
$\impairment = 0$. The latter means that the trial was effectively a
plain application of the \rd. Noteworthily, the given sender strategy
approximate a pure sender strategy that crisply partitions the state
space into non-convex sets. The irregular shape of the receiver
strategy shows clearly that the pictured strategy pair has not yet
reached a dynamically stable state under the \rd. Indeed, the trial
did not converge in our technical sense, but was obtained by stopping
after 200 rounds of iterated \rdd. In contrast, the outcome of a trial
with identical parameters, except that imprecision was $\impairment =
0.1$, is shown in Figure~\ref{fig:example_stratsB} had converged after
98 rounds of iterated \rdd. The sender strategy shows a smooth
blending from one ``category'' to the other, and the receiver's
interpretations are rather extended curves, peaking at a central point
in the relevant ``categories''. What these examples show is that (i)
(speaker) strategies can approach non-convex strategies and that (ii)
inclusion of imprecision can lead to seemingly well-behaved, yet vague
strategies in the sense that we are after.

To explore further what happened in all of our 50 trials for each
parameter triple, we defined a small number of metrics that aim to
capture numerically how vague, communicatively efficient and generally
well-structured the resulting strategy pairs were.

\paragraph{Entropy.} This classic information-theoretic metric captures
the amount of uncertainty in a signaling strategy.  The most natural
way to define it is in terms of mixed strategies rather than
behavioral strategies.  Let us recall that mixed sender (receiver)
strategies are functions $\Smixed \in \Delta(\Messgs^\States)$
($\Rmixed \in (\Acts^\Messgs)$).  We can define the entropy of a
sender and receiver strategy as
\begin{align*}
  E(\Smixed) & = \sum_{\Spure \in \Messgs^\States} \Smixed(\Spure) \cdot
  \log(\Smixed(\Spure)) & 
  E(\Rmixed) & = \sum_{\Rpure \in \Acts^\Messgs} \Rmixed(\Rpure) \cdot \log(\Rmixed(\Rpure)) \,.
\end{align*} 
These metrics are computationally expensive to calculate, since the
size of the domain over which the sum is computed grows exponentially
with the number of choice points.  Therefore, we
converted\footnote{See Appendix~\ref{sec:proofs}.}\todo{include this!?} the above
definitions into equivalent metrics defined in terms of behavioral
strategies.  These are as follows:
\begin{align*}
  E(\Sstrat) = -\sum_{\state \in \States} \sum_{\messg \in \Messgs}
  \Sstrat(\messg \probbar \state) \cdot \log(\Sstrat(\messg \probbar
  \state)) 
\end{align*} 
\begin{align*}
  E(\Rstrat) = -\sum_{\messg \in \Messgs} \sum_{\act \in \Acts}
  \Rstrat(\act \probbar \messg) \cdot \log(\Rstrat(\act \probbar
  \messg)) \,. 
\end{align*}
The metrics are lower bounded by $0$ and upper bounded by,
respectively, $\log(|\Messgs^\States|) = |\States| \cdot
\log(|\Messgs|)$ and $\log(|\Acts^\Messgs|) = |\Messgs| \cdot
\log(|\Acts|)$, thus we can normalize them by dividing by these
values. The reported values are these normalized values.

\paragraph{Informativity.} The quantity of information in a signal is
an old notion that also goes back to the start of information theory.
Skyrms~\citet[ch.~3]{Skyrms2010:Signals} discusses its use in the
context of signaling games.  The main idea is that we can quantify the
amount of information about a state $\state$ in a message $\messg$ via
the relation between the probability of the state given the message
$\Pr(\state \,|\, \messg)$ and the unconditional probability of the
state $\Pr(\state)$.  Following Bayes' theorem, we can express
$\Pr(\state \,|\, \messg)$ as $\frac{\Pr(\state) \cdot \Pr(\messg
  \,|\, \state)}{\Pr(\messg)}$.  We then have $\Pr(\messg \,|\,
\state) = \Sstrat(\state,\messg)$ and $\Pr(\messg) =
\sum_{\state^\prime \in \States} \Pr(\state^\prime) \cdot
\Sstrat(\state^\prime,\messg)$.  Finally, we equate sender
infomativity with the average overall information about states in each
signal.  Based on the definition by
Skyrms~\citet[p.~36]{Skyrms2010:Signals}, and the considerations
above, we define sender informativity as follows:
\begin{align*}
  I(\Sstrat) = \frac{1}{|\Messgs|} \sum_{\messg \in \Messgs} \sum_{\state \in \States} \frac{\Pr(\state) \cdot \Sstrat(\state,\messg)}{\sum_{\state^\prime \in \States} \Pr(\state^\prime) \cdot \Sstrat(\state^\prime,\messg)} \cdot \log \left(\frac{\Sstrat(\state,\messg)}{\sum_{\state^\prime \in \States} \Pr(\state^\prime) \cdot \Sstrat(\state^\prime,\messg)} \right) \,.
\end{align*}
Conversely, we can quantify the amount of information about an act in a message.
We equate receiver informativity with the average overall information about acts in each signal.
Based on Skyrms' definition~\citet[p.~39]{Skyrms2010:Signals}, we define receiver informativity as follows:
\begin{align*}
  I(\Sstrat,\Rstrat) = \frac{1}{|\Messgs|} \sum_{\messg \in \Messgs} \sum_{\act \in \Acts} \Rstrat(\messg,\act) \cdot \log \left(\frac{\Rstrat(\messg,\act)}{\sum_{\state \in \States} \Pr(\state) \cdot \sum_{\messg^\prime \in \Messgs} \Sstrat(\state,\messg^\prime) \cdot \Rstrat(\messg^\prime,\act)} \right) \,.
\end{align*}
Both metrics are bounded between $0$ and $1$.

@@

We run the noise-perturbed replicator dynamics with random initial strategies $\Sstrat_0$ and $\Rstrat_0$ until a convergence criterion is met, namely that the total absolute change in both sender and receiver strategies does not amount to more than $0.01$, \emph{i.e.}:
